---
title: "Lineage Definitions in ProVoC"
author: "Devan Becker"
bibliography: provoc_package.bib
execute:
  cache: true
format: html
---

## Introduction

This vignette demonstrates some different ways of defining and comparing lineage definitions, including a discussion of why it matters.

## Data Prep

This package contains data from @baaijensLineageAbundanceEstimation2022, as downloaded from NCBI SRA and processed via a custom pipeline (heavily based on GromStole, with some post-processing). 

In this paper, they explicitly look for B.1.1.7, B.1.427, B.1.429, and B.1.526. I'm going to add B.1.617.2 to this list, and take this list as given (another vignette will be written to test this assumption).

To make the code run faster, I'll look at every third day of data.

```{r}
#| label: get-data
#library(provoc)
devtools::load_all()
library(lubridate) # ymd()
library(here) # all paths relative to .git folder
data(Baaijens)

Baaijens$date <- ymd(Baaijens$date)
dates <- sort(unique(Baaijens$date))
dates3 <- (1:floor(length(dates) / 3)) * 3
dates <- dates[dates3]

b3 <- Baaijens[Baaijens$date %in% dates, 
    c("count", "coverage", "sra", "date", "label")]
b3$mutation <- parse_mutations(b3$label)
head(b3)
```

## Built-in Constellation Files

The PANGO team used to maintain a GitHub repository called "[Constellations](https://github.com/cov-lineages/constellations)" [@CovlineagesConstellations2023], which contained mutations that were believed to be more-or-less unique to the lineages. This is no longer updated, but is useful for historical analyses.

This set of lineage definitions is available via the `astronomize()` function.

```{r}
#| label: get-astro
astro <- astronomize() |>
    filter_lineages(c("B.1.1.7", "B.1.427", "B.1.429", "B.1.526", "B.1.617.2"))
dim(astro)
astro[, 1:5]
```

In this matrix, the rows represent the lineages and the columns represent the mutations, with a 1 representing the presence of a mutation in a given lineage.

Notice the use of `filter_lineages()`. This function gets the rows corresponding to the lineages, but it also ensures that we are only dealing with columns that have at least one "1". In other words, it only uses mutations that are present in at least one lineage. 

I have written a function to visualize this matrix:

```{r}
#| label: astro-defs
plot_lineage_defs(astro)
```

Only the shared mutations are shown, with the last column indicating how many unique mutations remain in each definition. Every other mutation gets a grey background to make it easier to follow a mutation when the matrices are larger. 

## Usher Barcodes

The [Freyja](https://github.com/andersen-lab/Freyja) [@karthikeyanWastewaterSequencingReveals2022; @AndersenlabFreyja2024] repository keeps an up-to-date barcodes file, which serves the same purpose as the definitions in constellations.

The `usher_barcodes()` function will check for this file, then download it if it's not present in a number of common locations. It's quite a large file, so I recommend storing the file and using the `filter_lineages()` function.

```{r}
#| label: get-bar
bar <- provoc::usher_barcodes(path = "working") |>
    filter_lineages(c("B.1.1.7", "B.1.427", "B.1.429", "B.1.526", "B.1.617.2"))
dim(bar)
```

## Constellations versus Barcodes

Let's quickly look at the mutations present in `astro` but not `bar`, then the mutations present in `bar` but not `astro`, as well as mutations that are in both.

```{r}
#| label: def-compare-setdiff
plot_lineage_defs2(astro, bar,
    main = "Constellations (red) versus Usher Barcodes (blue)")
```

From this plot, we can see that there are quite a few mutations defined in `astro` but not in `bar`. The blue bars indicate that the mutation is used in the lineage definition in `bar` but not `astro`; there are four mutations defined in both B.1.1.7 and B.1.617.2 by usher barcodes that are not present in the constellations file, and two of those mutations were also present in B.1.429. The barcodes seem to be a little bit more lenient about what mutations to include; constellations tried to get more-or-less "unique" mutations. 

## Results

Since we chose lineages when choosing the lineage definitions, we can use the convenient "~ ." notation to fit the models with all lineages.

Note that the Baaijens data were processed in such a way that each sample contains the same mutations. This means that each bar you see below is based on the exact same set of mutations. This isn't necessary, but I believe it improves the interpretability of the results.

```{r}
#| label: gg-lapis-bar
library(ggplot2)
library(patchwork) # Patching ggplots together
res_astro <- provoc(count / coverage ~ .,
    lineage_defs = astro,
    data = b3, by = "sra")
res_astro$date <- ymd(res_astro$date)
res_bar <- provoc(count / coverage ~ .,
    lineage_defs = bar,
    data = b3, by = "sra")
res_bar$date <- ymd(res_bar$date)

gg_astro <- autoplot(res_astro, date_col = "date") +
    labs(title = "Constellations")
gg_bar <- autoplot(res_bar, date_col = "date") +
    labs(title = "Usher Barcodes")

gg_astro / gg_bar + plot_layout(guides = "collect") &
    theme_bw() &
    facet_wrap(~ lineage, nrow = 1)
```

Those are two *very* different looking plots! We saw before that the lineage definitions are different, but let's check how they were actually used in the data. The same function can be used from before, which extracts the lineage definitions that were actually used (ignoring any mutations that were part of the definition but not present in the data).

```{r}
#| label: plot_lineage_defs2_with_data
plot_lineage_defs2(res_astro, res_bar,
    main = "Constellations (red) versus Usher Barcodes (blue)")
```

From this result, we can see that constellations only used four mutations to define B.1.617.2! The barcodes used 6! This is because there were only 4 (or 6) mutations in B.1.617.2 that were also present in the data. Let's dig a little deeper into this.

```{r}
#| label: plot-actual-defs
plot_actual_defs(res_astro, main = "Constellations")
plot_actual_defs(res_bar, main = "Usher Barcodes")
```

The two different matrices have very different definitions of B.1.617.2, and this explains the difference in results.

Furthermore, B.1.427 and B.1.429 had important differences - the barcodes have more mutations that overlap, meaning that the "de-mixing" algorithm must choose between one or the other. I'll expand on this in a "Choosing Lineages to Search For" vignette.

## LAPIS (CoV-Spectrum)

Since I have not managed to find where to get an API key, the mutations within each lineage were manually downloaded from cov-spectrum.org [@chenCoVSpectrumAnalysisGlobally2022a]. I set the date range to go from 2020-01-01 to 2021-04-27, with the latter being the last sample in the Baaijens data. 

The following code is very very hopefully going to change very soon. An API key would be nice, and I desperately need better mutation processing. It is, however, comforting that the column names match with `bar`, so at least my parsing is consistent!

```{r}
#| label: get-lapis
lineages <- c("B.1.1.7", "B.1.427", "B.1.429", "B.1.526", "B.1.617.2")
lapis_list <- lapply(lineages, function(x) {
    file <- read.csv(here("working", paste0(x, ".csv")))
    file$mutation[file$proportion > 0.8]
})
names(lapis_list) <- lineages
lapis <- lineage_defs_from_list(lapis_list)

# Avoiding parsing deletions, only looking at substitutions
lapis <- lapis[, !grepl("-", colnames(lapis))]
# TODO: user-friendly parsing
colnames(lapis) <- sapply(colnames(lapis), function(x) {
    provoc:::parse_mutation(
        type = "~",
        pos = substr(x, 2, nchar(x) - 1),
        alt = substr(x, nchar(x), nchar(x)))
})

plot_lineage_defs(lapis)
```

This looks a little different from before. There are 4 mutations that seem to be in every lineage, which doesn't give the de-mixing algorithm any information about which lineages are present. B.1.1.7 seems to have a lot of unique mutations, which might make the results more accurate (if the mutations truly are unique to B.1.1.7).

Let's quickly check how this compares to the barcodes:

```{r}
plot_lineage_defs2(bar, astro)
```

There are a couple mutations in astro but not bar, and a couple in bar but not astro. This is to be expected. However, those red bars in the shared mutation section are noteworthy! They are the same four mutations that we saw in the plot before, which told us that the de-mixing algorithm will have a hard time differentiating the lineages. Let's check the results.

```{r}
#| label: gg-astro-bar-lapis
#| fig-height: 5
res_lapis <- provoc(count / coverage ~ .,
    lineage_defs = lapis,
    data = b3, by = "sra")
res_lapis$date <- ymd(res_lapis$date)

gg_lapis <- autoplot(res_lapis, date_col = "date") +
    labs(title = "LAPIS (CoV-Spectrum)")

gg_astro / gg_bar / gg_lapis + plot_layout(guides = "collect") &
    theme_bw() &
    facet_grid(~ lineage)
```

Usher Barcodes and Cov-Spectrum generally agree, but what's going on with B.1.617.2??? It sort of looks like B.1.429 has taken some of the weight from B.1.617.2, but this is worth looking into. For example, did B.1.429 have mutations that are in B.1.617.2? From the comparison plot, there appear to be a lot of mutations that differentiate B.1.429 and B.1.617.2, so that can't be it. Maybe it's the coverage?

```{r}
#| label: lapis-coverage
plot_actual_defs(res_lapis)
```

It doesn't look like the mutations that differentiate B.1.617.2 and B.1.429 had low coverage. The minimum of the average coverages across samples^[That is, for each mutation and lineage, I found the average coverage across samples. The numbers on the plot below the lineage names represent the minimum, median, and maximum of the coverages shown in the plot.] isn't low for the shared mutations.

I think the primary conclusion here is that the difference in mutations between the barcodes and the lapis definitions is important enough to the demixing algorithm to give the differences in results. Perhaps the next plotting function I make should show something about the frequencies of the mutations in each lineage.

# Conclusion

Hopefully this vignette has successfully completed its two goals:

1. Demonstrate some of the functionality of `provoc` with lineage definitions.
2. Demonstrate the importance of choosing good lineage definitions!


# References
